---
title: "Installing Simulation Software"
author: "Jinko Graham"
date: "26/07/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

**NOTE 1** This document installs software. It is not meant to be "knitted",
but rather run line-by-line to install software on your computer.

**NOTE 2** Before working through this installation document, you 
must copy, or "clone" the class GitHub repository to your computer.
The simplest way to do this is to create an RStudio project linked to the git
repository https://github.com/SFUStatgen/SFUStat840. See the instructions
at https://support.rstudio.com/hc/en-us/articles/200532077-Version-Control-with-Git-and-SVN for getting started with git and version control in RStudio. Pay particular attention to the section "Creating a new project based on a remote Git or Subversion repository". An alternative method for cloning the class repository
that works on Mac and Linux is to use the command-line version of git, 
by typing `git clone https://github.com/SFUStatgen/SFUStat840` into a terminal.



## Overview

We will work with simulation software called `msprime` 
and `SLiM`. `msprime` simulates ancestries of chromosomes
according to a backward-in-time model called the coalescent 
that we will study in class. `SLiM` simulates genomic data
forwards-in-time using a model called the Wright-Fisher that 
we will also study in class. Both programs return a data structure
called a succinct tree sequence (STS)
that represent ancestral relationships between chromosomes.
`msprime` is a Python package (a.k.a. "module") and `SLiM` is a C program.
Installation of `msprime` also installs a toolkit called `tskit`
that is useful manipulation of STSs in Python.
A Python package called `pyslim` allows us to read STSs
output by `SLiM` into Python so that we can manipulation them
with the tools in `tskit`.
The majority of this document is comprised of instructions
for installing Python and the necessary package. At the
end is a short section describing installation of `SLiM`.

## Python and Python packages

I will be using an R package called **reticulate**
to interface with Python and the Python packages. 
The following instructions will lead you through
the installation of Python, reticulate `msprime`, `tskit` and `pyslim`.
We will also install a package called `tsinfer` that infers STSs from
genetic sequence data. 

### Install Python and conda

Conda is a 
package management system used by the Python community. Python and conda 
are installed simultaneously by a program called miniconda, which you can find at
  https://docs.conda.io/en/latest/miniconda.html
  
Choose the installer for Python version 3.7 that is 
appropriate for your operating system and follow the instructions 
to install.
Note for Mac users: You can choose either 
a "bash installer" or a ".pkg installer". I think the .pkg installer
will be more familiar to most users. Download the .pkg file to 
your computer,
double-click to start the installer and 
you will be guided through the installation.
    * NB: The installation of conda changes the configuration of the shell 
    used in your terminal windows. You need to open a *new* terminal
    after the installation to be able to use conda. From a new terminal,
    type `which python` and record the path to the python executable 
    that was installed by miniconda (you will need it in the next step). 
    On my Mac the path is
    `/Users/jgraham/miniconda3/bin/python`.


### Install reticulate

This is an R package and can be installed from CRAN in the usual way.
After installing, use `library()` to load the package and the function
`use_python()` to set the path to your Python executable:


```{r}
library(reticulate)
use_python("~/miniconda3/bin/python")
```

### Installing `msprime` and `tskit` into a Python "environment".

Python environments are like RStudio projects in that they are meant to
encapsulte all code and data for a specific project. However, they take
this encapsulation idea a
step further and also include any Python package
that you need for the project. In the following I create an environment called
`Stat840` with the `conda_create()` function
and install `msprime` and `tskit` into it with `conda_install()`.
Since `msprime` depends on `tskit` (in much the same way as 
R packages can depend on each other), `tskit` will be automatically
installed when we install `msprime`.

```{r}
envname <- "Stat840"
conda_create(envname)
conda_install(envname,"msprime")
```

Now try executing the following Python code chunk to test your installation
of `msprime`. (We'll learn more about the simulation parameters and
what the output means in class.)

```{python}
import msprime
ts = msprime.simulate(sample_size=4, recombination_rate=1e-8, Ne=1000)
tree = ts.first()
print(tree.total_branch_length)
```

### Installing `pyslim`

Unfortunately `pyslim` has not yet been made available on the
`conda-forge` website that hosts conda-installable packages. I
built a conda-installable version of `pyslim` myself and uploaded
it to the class GitHub site. I've also written an function 
`conda_install_local()` to simplify the installation of pyslim
from within R. Both the package and R function are in the 
class GitHub repository (https://github.com/SFUStatgen/SFUStat840). 
Before proceeding, you **must**
have a copy of the repository on your computer. 
In what follows, replace `/path/to/SFUStat840` with the path to 
the copy of the class repository on your computer.

```{r}
pkgname <- "pyslim"
dir <- "/path/to/SFUStat840/PythonPackages/" 
# NOTE to JG: the following points to a temporary location for the class 
# repository. Change when you decide where to put the class repository.
dir <- "/Users/jgraham/Dropbox/StatgenAdmin/Scratch840/SFUStat840/PythonPackages/"
source(paste0(dir,"conda_install_local.R"))
conda_install_local(envname,pkgname,dir)
```

Testing of `pyslim` is deferred until we've installed SLiM (see below).

### Installing `tsinfer`

`tsinfer` is also not available on conda-forge. I was not able to build
a conda-installable version of this package, so we must use
the Python package manager called `pip` that installs packages
from a repository called PyPI. The `conda_install()` function
can install packages *via* pip with the `pip=TRUE` argument. 
This also installs all of the package dependencies of `tsinfer`.
One of these dependencies is
called `numcodecs`, and the pip-installed version of `numcodecs` is
broken. The work-around is to re-install `numcodecs` with conda,
after installing `tsinfer`.

```{r}
conda_install("Stat840","tsinfer",pip=TRUE)
conda_install("Stat840","numcodecs")
```

You can test `tsinfer` with the following Python code chunk:

```{python}
import tsinfer
with tsinfer.SampleData(sequence_length=4) as sample_data:
    sample_data.add_site(0, [0, 1, 0, 0, 0], ["A", "T"])
    sample_data.add_site(1, [0, 0, 0, 1, 1], ["G", "C"])
    sample_data.add_site(2, [0, 1, 1, 0, 0], ["C", "A"])
    sample_data.add_site(3, [0, 1, 1, 0, 0], ["G", "C"])
inferred_ts = tsinfer.infer(sample_data)
tree = inferred_ts.first()
print(tree.total_branch_length)
```

### Using the Stat840 Python environment in future R sessions

To use the `Stat840` Python environment and all the package
you have installed into it, include the following code 
chunk at the top of your RMarkdown documents.

```{r}
library(reticulate)
use_condaenv("Stat840")
```


## Installing `SLiM`

`SLiM` is a C program available from 
  https://messerlab.org/slim/
  
Unfortunately the program is only available for Mac OS X and Linux users.
In fact, our group has only been successful installing SLiM on Mac OS X.
If you are a Mac user, download and run the OS X installer from the 
above website. 

If you do not have access to a Mac, you can get an 
account on a Mac Pro used by the Graham and McNeney labs. The computer
is housed in Brad McNeney's office, and you will have to make an 
appointment with Brad to log in from the console the first time you
use it. After the initial login, you can log in remotely. You will
not be able to use the SLiM GUI, or any other GUI, when you log in remotely.
Also, without GUI access you can't use RStudio, so 
you will need to use `git` from the 
command line to clone the repository to your account on this Mac;
i.e., run `git clone https://github.com/SFUStatgen/SFUStat840` .

After installing SLiM on your computer, or getting access to the G&M Mac, you
can test it by typing the following into a
terminal window:

```
slim /path/to/SFUStat840/SLiM/HallerExample.txt
```

(This is a modified example from the paper by Haller et al. (2018)  that
introduces SLiM version 3 and its capability to store simulation results
as succinct tree sequences.)
After running SLiM you should find a file called `ex1_TS.trees` in your working directory. A copy of this file is also on the class GitHub repository
in the SFUStat840/SLiM folder.

Now that you have a file of SLiM output, you can also 
test your installation of `pyslim`  as follows:

```{python}
import pyslim
#ts = pyslim.load("/path/to/SFUStat840/SLiM/ex1_TS.trees")
ts = pyslim.load("/Users/jgraham/Dropbox/StatgenAdmin/Scratch840/SFUStat840/SLiM/ex1_TS.trees")
tree = ts.first()
print(tree.total_branch_length)
```

