---
title: "Test your Installed Software"
author: "Jinko Graham"
date: "30/07/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


If you successfully installed all the software 
in the `installSimSoftware.Rmd` document, you 
should be able to knit this document without 
errors. 
* Note 1: Before you knit, change the `envname`
variable in the following code chunk to the 
name of your Stat 840 conda environment, 
if different from "Stat840" and change
the character string `pathToSFUStat840`
to the location of the SFUStat840 repository on 
your computer.
* Note 2: The "test" of SLiM is not executed
in this document. Instead, you are given 
a command that you can type into a Mac
terminal to test that SLiM is 
installed properly.


```{r}
envname <- "Stat840"
pathToSFUStat840 <- "/Users/jgraham/Desktop/Drawer/TeachSemAdmin/Stat890_19/SFUStat840"
pathToSFUStat840 <- "/Users/mcneney/Dropbox/StatgenAdmin/Scratch840/SFUStat840"
```


## Test reticulate

```{r}
library(reticulate)
use_python("~/miniconda3/bin/python")
use_condaenv(envname)
```

## Test `msprime` and `tskit`

```{python}
import msprime
ts = msprime.simulate(sample_size=4, recombination_rate=1e-8, Ne=1000)
tree = ts.first()
print(tree.total_branch_length)
```


## Test `pyslim`

In the following Python code chunk, replace the path to the 
class repository on my computer (/Users/jgraham/ ... /SFUStat840) 
with the path on your computer.

```{python}
import pyslim
ts = pyslim.load("/Users/jgraham/Desktop/Drawer/TeachSemAdmin/Stat890_19/SFUStat840/SLiM/ex1_TS.trees")
tree = ts.first()
print(tree.total_branch_length)
```


## Test `SLiM`

After installing SLiM on your Mac, or getting access to the Graham and 
McNeney lab Mac, you
can test it by typing the following into a
terminal window.
Replace the path to the 
class repository on my computer (/Users/jgraham/ ... /SFUStat840) 
with the path on your computer.

```
slim /Users/jgraham/Desktop/Drawer/TeachSemAdmin/Stat890_19/SFUStat840/SLiM/HallerExample.txt
```

After running SLiM you should find a file called `ex1_TS.trees` in your working directory. A copy of this file is also on the class GitHub repository
in the SFUStat840/SLiM folder.
